---
title: "Regression Project - Analysis of Forest Fires"
author: "Shravan Kuchkula"
date: "9/20/2017"
output:
  github_document:
    toc: yes
  html_document:
    keep_md: yes
    theme: cosmo
    toc: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
---

## Introduction
Our main aim is to understand how the burned area of forest fires, in the northeast region of Portugal, is related to the meteorological and other data. The [*forest fires*](http://archive.ics.uci.edu/ml/datasets/Forest+Fires) data set is used to perform this analysis. The data set contains 13 attributes. A brief description of the variables is given here:

* **X** - x-axis spatial coordinate within the Montesinho park map: 1 to 9.
* **Y** - y-axis spatial coordinate within the Montesinho park map: 2 to 9.
* **month** - month of the year: 'jan' to 'dec'.
* **day** - day of the week: 'mon' to 'sun'.
* **FFMC** - FFMC index from the FWI system.
* **DMC** - DMC index from the FWI system.
* **DC** - DC index from the FWI system.
* **ISI** - ISI index from the FWI system.
* **temp** - temperature in Celsius degrees.
* **RH** - relative humidity in %.
* **wind** - wind speed in km/h.
* **rain** - outside rain in mm/m2.
* **area** - the burned area of the forest (in ha).

## Data gathering and cleaning
All the required libraries are loaded.

```{r message = FALSE, warning = FALSE}
source('libraries.R')
```

The data is retrieved and stored into a data frame using `read.csv`.

```{r}
forestfires <- read.csv("http://archive.ics.uci.edu/ml/machine-learning-databases/forest-fires/forestfires.csv",
                        stringsAsFactors = FALSE)
```


## Exploratory Data Analysis
The response variable in our analysis is the `area`. A histogram of `area` reveals that a large number of values are zeros. A zero value for area indicates that there were no forest fires. 

```{r message=FALSE , warning=FALSE}
attach(forestfires)
hist(area, 40)
```

A log transformation can reveal the distribution of the area variable. However, since we have zeros, the log transform should be done on area + 1, i.e, log(area + 1). The distribution of log(area + 1) is shown:

```{r message=FALSE, warning=FALSE}
attach(forestfires)
hist(log(area + 1), 40)
```

Even this log(area + 1) transformation does not help much as large number of values are zeros. A quick check reveals that 47.77 % of the time, there has been no forest fire. 

```{r}
sum(forestfires$area == 0)/nrow(forestfires)
```

Since our aim is to predict the area of the forest fires given that there is indeed a forest fire going on, and not to predict whether there will be a forest fire or not, we can remove the observations with area = 0 from our analysis. 

The histogram for log(area[area >0]) is shown here:
```{r message=FALSE, warning=FALSE}
attach(forestfires)
hist(log(area[area>0]), 40)
```

Next, we will see how the transformed response variable correlates with other variables. X, Y, month and day are categrorical variables. A better way to explore the relationship of these categorical variables with the log(area[area>0]) is to draw a box plot. 

```{r message=FALSE, warning=FALSE}
attach(forestfires)
forestfires %>%
  filter(area > 0) %>%
  mutate(logArea = log(area)) %>%
  ggplot(aes(x = as.factor(X), y=logArea)) +
  geom_boxplot()
```

The box plot reveals that the mean and variation is pretty much the same across all levels of X, which indicates that X may not be a great predictor of logArea.

```{r message=FALSE, warning=FALSE}
attach(forestfires)
forestfires %>%
  filter(area > 0) %>%
  mutate(logArea = log(area)) %>%
  ggplot(aes(x = as.factor(Y), y=logArea)) +
  geom_boxplot()
```

Create a correlation matrix of all the numeric variables:

```{r message=FALSE, warning=FALSE}
attach(forestfires)
numericFF <- forestfires %>%
  filter(area > 0) %>%
  mutate(logArea = log(area)) %>%
  select(-X, -Y, -month, -day, -area)

# Create correlation matrix
round(cor(numericFF), 2)
```

Visualize this in a corrplot

```{r message=FALSE, warning=FALSE}
attach(forestfires)
numericFF <- forestfires %>%
  filter(area > 0) %>%
  mutate(logArea = log(area)) %>%
  select(-X, -Y, -month, -day, -area)

# Create correlation matrix
M <- round(cor(numericFF), 2)
corrplot(M, method="pie", type="lower")
```

There appear to be some correlation between predictor variables. Very little correlation with the response variable.


  